Proofs as Programs in Classical Logic\\Notes

### 2022-02-13 Notes/notes.tex
Code
implementation

\usepackage{amsmath,amssymb,amsthm}
\usepackage{ wasysym }
\usepackage{ stmaryrd }
\usepackage{ mathpartir }
\usepackage{xcolor}
\usepackage{bussproofs}
\usepackage[left=3.00cm, right=3.00cm, top=3.00cm,bottom=3.0cm]{geometry}

@@

>>>>

\usepackage{amsmath,amssymb,amsthm}
\usepackage{ wasysym }
\usepackage{ stmaryrd }
\usepackage{ mathpartir }
\usepackage{xcolor}
\usepackage{bussproofs}
\usepackage[left=2.00cm, right=2.00cm, top=2.00cm,bottom=2.0cm]{geometry}
\usepackage{pdflscape}

@@

To explain our approach let us first recall how resolution works and why it is effective for classical logic. In a nutshell the principle of resolution works as follows: From a set of formulas $A, B, C\dots$ obtain a set of sequents $A'_1\dots  A'_n\vdash A_1\dots A_n$, $B'_1\dots B'_n\vdash B_1\dots B_n$, $C'_1\dots C'_n\vdash C_1\dots C_n$ in which all occurring formulas are somehow simple. Then iteratively apply the resolution rule (and some other rules, but this is the crucial one), which is an analogon to cut, i.e.
\begin{center}
	\AxiomC{$A, \Gamma\vdash \Delta$}
	\AxiomC{$\Gamma'\vdash A', \Delta'$}
	\BinaryInfC{$(\Gamma\cup \Gamma')\theta\vdash (\Delta\cup\Delta')\theta$}
	\DisplayProof
\end{center}
where $\theta$ is the mgu of $A$ and $A'$, and by process of saturation (attempt to) obtain the empty Sequent.

Now for classical logic we can simply interpret $A'_1,\dots  A'_n\vdash A_1,\dots, A_n$ as $\neg A'_1\vee\dots\vee\neg A'_n\vee A_1\vee\dots\vee A_n$. Furthermore we can normalize every formula such that each of the $A'_i, A_i$ is atomic. The resolution calculus is then immediately refutationally complete, i.e. if $A, B, C\dots $ are inconsistent there exists a successful resolution, giving a refutation of $A\wedge B\wedge C\dots$, which is classically equivalent to a proof of $B\wedge C\dots\rightarrow\neg A$. I.e. if we want to prove $B\wedge C\dots\rightarrow A$ we can simply preform resolution on $\neg A\wedge B\wedge C\dots$

Now there are a number of hurdles in applying this strategy to intuitionistic logic, or to extract intuitionistic proof from classical resolution proofs. First and foremost is the non-existence of normal forms. In particular with the (intuitionally speaking) strict interpretation of $A'_1,\dots  A'_n\vdash A_1,\dots, A_n$ as $\neg A'_1\vee\dots\vee\neg A'_n\vee A_1\vee\dots\vee A_n$ we can hardly translate any formulas into such a form where the $A_i$ are atomic, or even literals. This can be remedied of course by interpreting $A'_1,\dots  A'_n\vdash A_1,\dots, A_n$ as $\neg(A'_1\wedge\dots\wedge A'_n\wedge(\neg A_1\vee\dots\vee \neg A_n))$, i.e. a double negation translation. But this brings us to the second problem: Even if our calculus is refutationally complete a refutation of $\neg A\wedge B\wedge C\dots$ does not give us a proof of $B\wedge C\dots\to A$ but rather $B\wedge C\dots\to \neg\neg A$. This problem is especially pronounced in the second interpretation of $\vdash$ where we are not even using $A$ but $\neg \neg A$.

Now there are a few saving graces: First of all, if $A$ is decidable, i.e. $\neg\neg A\to A$, we immediately obtain a proof of $A$. Furthermore we may attempt Friedman Translation~\cite{Friedman_1978} of our proof and indeed~\cite{Berger_2002} gives us some quite liberal criteria when this is possible in principle, e.g. when $B, C\dots$ are theorems of PA and $A$ is a $\Pi_0^2$ formula. And indeed interpreting $A'_1,\dots  A'_n\vdash A_1,\dots, A_n$ as $A'_1\to\dots\to A'_n\to A_1\vee\dots\vee A_n$, we are able to obtain an intuitionistic proof if their conditions are met and the input formulas can be transformed to such normal forms.
